---
title: "Russian_Olive_Modeling"
author: "Tobin Haefele"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
library(maxnet,warn.conflicts = FALSE)
library(readxl)
library(dplyr)
library(sf)
library(raster)
library(geodata)
library(terra)
library(corrplot)
library(arcgisbinding)

knitr::opts_chunk$set(echo = TRUE)
```

## Import data

```{r data import, warning=FALSE}
#import observations
survey.df <- read_excel("data/MTNHP_RussianOlive.xlsx", 
    sheet = "Observations")

#import climate data 
climate_data <- geodata::worldclim_tile(11, 12, var = "bio", res = 0.5, path = "data/") #not sure the tile is correct

#import montana shapefile
mt_shape <- st_read("data/MontanaStateBoundary_shp/StateofMontana.shp")

```

## Cleaning data
```{r data cleaning, echo=FALSE}

#remove spaces from column names
colnames(survey.df) <- gsub(" ", "_", colnames(survey.df))

#rename spatial precision column for easier access
colnames(survey.df)[colnames(survey.df) == "Spatial_Precision_(meters)"] <- "Spatial_Precision"

#transform montana shapefile to crs
mt_shape <- st_transform(mt_shape, crs = 4326)

# Crop the climate data to the extent of the Montana shapefile
climate_data_cropped <- crop(climate_data, mt_shape)

# Mask the cropped climate data to the Montana state boundary
climate_data_masked <- mask(climate_data_cropped, mt_shape)

#remove observations where spatial precision is > 800 
survey.df.filtered <- survey.df[survey.df$Spatial_Precision< 800,]

```

## Testing cleaned data
```{r plot, echo=FALSE}

#create a small data frame with the lat and lon of the observations
df <- data.frame(
  "Lat" = survey.df.filtered$Latitude,
  "Lon" = survey.df.filtered$Longitude
)

#convert to sf object
df <- st_as_sf(df, coords = c("Lon", "Lat"), crs = 4326)

#plot the points
plot(climate_data_masked[[3]], main = "Annual Mean Temperature in Montana")
plot(st_geometry(mt_shape), add = TRUE, border = "red")
plot(df, add = TRUE, pch = 20, col = "red")


```

## Prepping the data for modeling
```{r model prep, echo=FALSE}
set.seed(1000)

#Create pseudo-absence points within the Montana boundary
pseudo_absence <- st_sample(mt_shape, size = (nrow(df)*2), type = "random") %>% 
  st_as_sf()

#add pa column to pseudo-absence points and df
df <- df %>% mutate(pa = 1)
pseudo_absence <- pseudo_absence %>% mutate(pa = 0)

pseudo_absence <- pseudo_absence %>%
  rename(geometry = x)

#combine the two dataframes
combined_df <- rbind(df,pseudo_absence)

#extract the climate data values for the points
climate_values <- extract(climate_data_masked, combined_df)

#loop through climatte values and rename all columns after the first
for(i in 2:20){
  colnames(climate_values)[i] <- paste("bio", i-1, sep = "")
}

#combine the climate values with the points
combined_df <- cbind(combined_df, climate_values)
```

## Basic modeling
```{r model, echo=FALSE}
# create correlation matrix
df.cor.climate <- st_drop_geometry(combined_df)
df.cor.climate <- df.cor.climate[3:21]
df.cor.climate <- na.omit(df.cor.climate)
df.cor.climate <- df.cor.climate[, sapply(df.cor.climate, is.numeric)]
climate.cor <- cor(df.cor.climate)

#view the correlation matrix
corrplot(climate.cor, method = "number")

#create test model
mod1.LR <- glm(as.factor(pa) ~ bio1 + bio2 + bio3 + bio4 + bio5 + bio6 + bio7 + bio8 + bio9 + bio10 + bio11 + bio12 + bio13 + bio14 + bio15 + bio16 + bio17 + bio18 + bio19, data = combined_df, family = binomial)

#predict the model
mod1.pred <- predict(mod1.LR, type = "response")

#summarize the model
summary(mod1.LR)

```

### Potential Variables to include
- Wetland Riparian (land cover )
- Maximum Summer Temp (climate)
- Frost Free Days (climate)
-  Anthropogenic Influence (land cover)
- Introduced Vegetation (land cover)
- Soil pH (soil)
- Forest - Conifer (land cover)
- Bulk Density (soil)
- Degree Days (climate)
-  Distance to Water Edge (hydrography)

### Data sets 
https://geoinfo.msl.mt.gov/msdi/land_use_land_cover
https://geoinfo.msl.mt.gov/msdi/climate/
https://prism.oregonstate.edu/
https://doi.org/10.18113/S1KW2H (soil)